---
title: Emailing Files with SendmailR and Purrr
author: Jacob Schwan
date: '2018-02-12'
slug: sendmailr-purrr
categories: ["R"]
tags: ["sendmailr","purrr"]
banner: "banners/sendmailR_purrr.png"
images: [banners/sendmailR_purrr.png]
---

I use R in my day job to generate and send out regular reports. The [`sendmailR`](https://cran.r-project.org/web/packages/sendmailR/index.html) package has been great to automatically email an output file. However, it becomes increasingly annoying to attach multiple files to an email. Or, at least, it was before I started learning the [`purrr`](http://purrr.tidyverse.org/) package.

<!--more-->

## TL;DR

Here's the code template I use to attach all of the files in a specific folder to an email.

```{r, eval = FALSE}
library(sendmailR)
folder_name <- ""
email_from <- "<>"
email_to <- c("<>", "<>")
email_bcc <- "<>"
email_subject <- ""
email_body <- ""

output_files <- data.frame(filename = list.files(folder_name),
                           file_path = list.files(folder_name, full.names = T),
                           stringsAsFactors = F)

email_attachments <- purrr::map2(output_files$file_path,
                                 output_files$filename,
                                 mime_part)

sendmail(from = email_from,
         to = email_to,
         bcc = email_bcc,
         subject = email_subject,
         msg = c(email_body, email_attachments),
         control = list(smtpServer = ""))
```

## Motivation

As I mentioned above, attaching more than one document with `sendmailR` can become very tedious. Each file is defined as a separate `mime_part()` object with a filename and path, then added to a list along with the message body.  So my typical block of code at the end of a script would look like this:

```{r, eval = F}
library(sendmailR)

#Build Email
email_from <- "<myemail@mycompany.com>"
email_to <- c("<recipient1@mycompany.com>", "<recipient2@mycompany.com>")
email_subject <- paste0("Some report as of", Sys.Date())
email_body <- paste0("Hello recipeints,\n\nAttached are the reports you need as of ", Sys.Date(), ".\n\nRegards,\nJacob Schwan")
attachment1 <- mime_part(x = paste0(folder, file_name1), name = file_name1)
attachment2 <- mime_part(x = paste0(folder, file_name2), name = file_name2)
attachment3 <- mime_part(x = paste0(folder, file_name3), name = file_name3)
attachment4 <- mime_part(x = paste0(folder, file_name4), name = file_name4)

sendmail(from = email_from, to = email_to, bcc = email_from, subject = email_subject, 
         msg = list(email_body, attachment1, attachment2, attachment3, attachment4),
         control = list(smtpServer = "smtp.mycompany.com"))
```

This works okay, but it's a lot of typing to define each attachment and makes it really easy to forget to build the `mime_part()` for a file or to add it to the `msg = list()` in the `sendmail()` function.  I would much prefer to just have the email send everything in a designated output folder. Then one day it clicked.  I want to dynamically make a list based on some input...that sounds like a job for the `purrr::map()` function!

## Building a Solution

I had read some introductions and tutorials on using `purrr` and the `map()` functions, and was already using `purrr::map_df(list.files(folder_name, pattern = "csv$", full.names = T), read_csv)` to read a whole folder of similar CSV files into a single data frame. So, one day I decided to take some time and figure out how to adapt this to email attachments. 

Reading the [`purrr`](http://purrr.tidyverse.org/) documentation made it pretty clear that `map2()` was the function I needed to work with.  It took some trial and error to figure out the best approach to building the input for the function.  I finally settled on building a data frame with the necessary file names and full paths.

```{r, eval = F}
output_files <- data.frame(filename = list.files(folder_name),
                           file_path = list.files(folder_name, full.names = T),
                           stringsAsFactors = F)
```

Then running the following `map2()` function would give me a nice list of `mime_part` objects.

```{r, eval = F}
email_attachments <- purrr::map2(output_files$file_path,
                                 output_files$filename,
                                 mime_part)
```

The hardest piece to figure out, surprisingly, was how to correctly combine this list with the body of my email in the `msg` argument of the `sendmail()` function.  I had originally thought that just using the `list()` function should work. But as you can see in this example, this creates a nested list which the `msg` argument can't handle.

```{r}
email_attachments <- list("Attachment1", "Attachement2", "Attachment3")
email_body <- "Body of email"

list(email_body, email_attachments)
```

After some frustrating time trying to Google a solution, I finally stumbled into it.  And it was surprisingly simple!

```{r}
c(email_body, email_attachments)
```

I was initially surpised when this worked, as I had always thought that `c()` was just for building vectors.  Turns out that the documentation for [`c()`]() states very clearly (in big bold letters at the top) that it can combine values into a vector *or list*.  Wish I had been told that years ago!

So now my standard email template looks like this:

```{r, eval = FALSE}
library(sendmailR)
folder_name <- ""
email_from <- "<>"
email_to <- c("<>", "<>")
email_bcc <- "<>"
email_subject <- ""
email_body <- ""

output_files <- data.frame(filename = list.files(folder_name),
                           file_path = list.files(folder_name, full.names = T),
                           stringsAsFactors = F)

email_attachments <- purrr::map2(output_files$file_path,
                                 output_files$filename,
                                 mime_part)

sendmail(from = email_from,
         to = email_to,
         bcc = email_bcc,
         subject = email_subject,
         msg = c(email_body, email_attachments),
         control = list(smtpServer = ""))
```

## What I Learned

Working though this problem really gave me some practice using lists.  Lists are one of my R weaknesses.  I don't feel like I've fully internalized when to use them and how to use them effectively.  Just learning to combine them with `c()` feels like a big win for me.

Questions, critiques, and feedback are welcome in the comments section below. My thanks in advance to anyone that takes the time to leave one.
